Probability Aggregation: Dynamic Hierarchical Modeling of Expert Beliefs in Geopolitical Events

### 2013-08-26 paper.tex
Edits (Second Draft)

\begin{abstract}
Most subjective probability aggregation procedures use a single probability judgement from each expert, even though it is common for experts studying real problems to update their probability estimates over time.  This paper makes an advance towards unexplored areas of probability aggregation by considering a dynamic context in which experts are allowed to update their beliefs at random intervals. The updates are assumed to occur very infrequently, resulting into a highly sparse dataset that cannot be modeled by standard time-series procedures. In response to the lack of appropriate methodology, this paper presents a dynamic model that takes into account the forecaster’s level of self-reported expertise and produces aggregate probabilities that are sharp and well-calibrated both in- and out-of-sample. The model is demonstrated on a real-world dataset that includes over 2,300 experts making multiple probability forecasts on different subsets of 166 different international political events. 
\end{abstract}

@@

This paper adds another layer of complexity by assuming a heterogeneous set of experts, most of whom only make one or two probability assessments over the hundred or so days before the event resolves. This means that the decision-maker faces a different group of experts  every day, with only a few experts returning later on for a second round of advice. The problem at hand is therefore strikingly different from many time-series estimation problems, where one has an observation at every time point -- or almost every time point. As a result, standard time-series procedures like ARIMA (see, e.g., \citet{mills1991time}) are not directly applicable to our setting.

In response to the lack of appropriate modeling procedures, this paper introduces a time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated estimate of the crowd belief. The model is highly interpretable and can be used for
\begin{itemize}
\item the analysis of group-level under- and overconfidence across different levels of self-reported expertise,
%\item confidence bands for the crowd belief,
\item accurate probability forecasts, and
\item many question-specific quantities that have easy interpretations and can be used to gain novel insight in the social sciences. 
\end{itemize}

@@

The data collection began with a recruitment of 2,365 forecasters ranging from graduate students to political science faculty and practitioners. The recruiting was made from professional societies, research centers, alumni associations, science bloggers, and word of mouth. Requirements included at least a Bachelor's degree and completion of psychological and political tests that took roughly two hours. These measures assessed cognitive styles, cognitive abilities, personality traits, political attitudes, and real-world knowledge. The experts were asked to give probability forecasts (to the second decimal point) and to self-assess their level of expertise (measured on a 1-to-5 scale with 1 = Not At All Expert and 5 = Extremely Expert) on a number of 166 geopolitical binary events taking place between September 29, 2011 and May 8, 2013. Each question was active for a period of time during which the participating experts were allowed to update their forecasts as frequently as they liked without being penalized. The forecasters knew that their probability estimates would be assessed for accuracy using Brier scores\footnote{The Brier score is the squared distance between the probability forecast and the event indicator that equals 1.0 or 0.0 depending on whether the event happened or not, respectively. See \citet{Brier} for the original introduction.}. This incentivized them to report their true beliefs instead of attempting to game the system (\citet{winkler1968good}). In addition to receiving \$150 for meeting minimum participation requirements that did not depend on prediction accuracy, the forecasters also received status rewards for their performance via leader-boards displaying Brier scores for the top 20 forecasters. Since a typical expert participated only in a small subset of the 166 questions, the forecasters are considered indistinguishable conditional on the self-reported level of expertise.
Tables \ref{DataStats} and \ref{ExpertiseTable} provide more relevant summary statistics on the data. Notice that the distribution of the self-reported expertise is skewed to the right and that some questions remained active longer than others.  For more details on the data and its collection see \citet{ungar2012good}.

@@

To illustrate the nature of the data with some concrete examples, Figures \ref{Example1} and \ref{Example2} show scatterplots of the probability forecasts given for (a) \textit{Will the expansion of the European bailout fund be ratified by all 17 Eurozone nations before 1 November 2011?}, and (b) \textit{Will the Nikkei 225 index finish trading at or above 9,500 on 30 September 2011?}. The points have been jittered slightly to make overlaps visible. The darkness of the points are positively associated with the self-reported expertise levels. Given that the European bailout fund was ratified before 1 November 2011 and that the Nikkei 225 index finished trading at around 8,700 on 30 September 2011, the general trend of the probability forecasts converge to the correct answers. There is, however, much disagreement among the experts. This disagreement persists even near the closing dates of the questions. Unfortunately, the data is not only noisy but also quite sparse. As mentioned before, the forecasters were allowed to update their forecast as long as the questions remained active. Updating was done on a very infrequent basis. An expert gave, on average, only 0.0159 forecasts per day, and so a typical expert made a forecast about every 62 days. As a result the average response rate was around 13.5 forecasts per day from a large group of experts.

@@

\section{Discussion}
This paper began with an introduction of a rather unorthodox but nonetheless realistic time-series setting in which probability forecasts are made very infrequently by a heterogeneous group of experts. The resulting data is too sparse to be modeled efficiently with standard time-series methods.  In response, our work introduces an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated crowd belief across time. The model is estimated in two-steps: The first step, known as the \textit{sampling step}, samples constrained versions of the model parameters via a Gibbs sampler that only makes use of standard distributions. The second step, known as the \textit{calibration step}, involves a one-dimensional optimization procedure that transforms the constrained parameter values to their unconstrained counterparts.

@@

\subsection{Limitations and Directions for Future Research}
This paper developed a time-series procedure that addresses the main challenges in modeling sparse probability forecasting data. The introduced model was kept parsimonious and can be viewed as a basic template that is amenable to many future extensions. For instance, most of the model parameters were assumed to remain constant over time. It is intuitively reasonable, however, that the behavior of the model parameters is different earlier as opposed to later on in the question.  For instance, the level of disagreement, $\sigma^2_k$, among the experts can be expected to decrease towards the final time point when the the question finally resolves. This hypothesis could be investigated by letting $\sigma^2_{t,k}$ depend on the previous term $\sigma^2_{t-1,k}$ similarly to the hidden process, $X_{t,k}$.

If personal information on the experts were inquired during the data collection process, it is possible to study the forecasting behavior of different kinds of experts. For instance, in this paper the multiplicative bias was modeled as a function of self-reported expertise. The model, however, is by no means restricted to the study of self-reported expertise. Parameter dependencies could be constructed based on many other expert characteristic, such as sex, education or specialty, to produce a range of novel insight on the forecasting behavior across different groups of experts. Furthermore, expert characteristics could be interacted with question types, such as economic, domestic or international, to learn about the kind of experts who generally perform well on certain types of questions. This would be of interest to the decision-maker who could use the information as a basis for consulting only a high-performing subset of the available experts.

>>>>

\begin{abstract}
Most subjective probability aggregation procedures use a single probability judgement from each expert, even though it is common for experts studying real problems to update their probability estimates over time.  This paper makes an advance towards unexplored areas of probability aggregation by considering a dynamic context in which experts are allowed to update their beliefs at random intervals. The updates are assumed to occur very infrequently, resulting into a highly sparse dataset that cannot be modeled by standard time-series procedures. In response to the lack of appropriate methodology, this paper presents a dynamic model that takes into account the forecaster's level of self-reported expertise and produces aggregate probabilities that are sharp and well-calibrated both in- and out-of-sample. The model is demonstrated on a real-world dataset that includes over 2,300 experts making multiple probability forecasts on different subsets of 166 different international political events. 
\end{abstract}

@@

This paper adds another layer of complexity by assuming a heterogeneous set of experts, most of whom only make one or two probability assessments over the hundred or so days before the event resolves. This means that the decision-maker faces a different group of experts  every day, with only a few experts returning later on for a second round of advice. The problem at hand is therefore strikingly different from many time-series estimation problems, where one has an observation at every time point -- or almost every time point. As a result, standard time-series procedures like ARIMA (see, e.g., \citet{mills1991time}) are not directly applicable to our setting. In response to the lack of appropriate modeling procedures, this paper introduces a time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated estimate of the crowd belief. The model is highly interpretable and can be used for
\begin{itemize}
\item the analysis of group-level under- and overconfidence across different levels of self-reported expertise,
%\item confidence bands for the crowd belief,
\item accurate probability forecasts, and
\item many question-specific quantities that have easy interpretations and can be used to gain novel insight in the social sciences. 
\end{itemize}

@@

The data collection began with recruitment of 2,365 forecasters ranging from graduate students to political science faculty and practitioners. The recruiting was made from professional societies, research centers, alumni associations, science bloggers, and word of mouth. Requirements included at least a Bachelor's degree and completion of psychological and political tests that took roughly two hours. These measures assessed cognitive styles, cognitive abilities, personality traits, political attitudes, and real-world knowledge. The experts were asked to give probability forecasts (to the second decimal point) and to self-assess their level of expertise (on a 1-to-5 scale with 1 = Not At All Expert and 5 = Extremely Expert) on a number of 166 geopolitical binary events taking place between September 29, 2011 and May 8, 2013. Each question was active for a period of time during which the participating experts were allowed to update their forecasts as frequently as they liked without being penalized. The forecasters knew that their probability estimates would be assessed for accuracy using Brier scores\footnote{The Brier score is the squared distance between the probability forecast and the event indicator that equals 1.0 or 0.0 depending on whether the event happened or not, respectively. See \citet{Brier} for the original introduction.}. This incentivized them to report their true beliefs instead of attempting to game the system (\citet{winkler1968good}). In addition to receiving \$150 for meeting minimum participation requirements that did not depend on prediction accuracy, the forecasters also received status rewards for their performance via leader-boards displaying Brier scores for the top 20 forecasters. Since a typical expert participated only in a small subset of the 166 questions, the forecasters are considered indistinguishable conditional on the level of self-reported expertise.
Tables \ref{DataStats} and \ref{ExpertiseTable} provide more relevant summary statistics on the data. Notice that the distribution of the self-reported expertise is skewed to the right and that some questions remained active longer than others.  For more details on the data and its collection see \citet{ungar2012good}.

@@

To illustrate the nature of the data with some concrete examples, Figures \ref{Example1} and \ref{Example2} show scatterplots of the probability forecasts given for (a) \textit{Will the expansion of the European bailout fund be ratified by all 17 Eurozone nations before 1 November 2011?}, and (b) \textit{Will the Nikkei 225 index finish trading at or above 9,500 on 30 September 2011?}. The points have been jittered slightly to make overlaps visible. The darkness of the points is positively associated with the self-reported expertise. Given that the European bailout fund was ratified before 1 November 2011 and that the Nikkei 225 index finished trading at around 8,700 on 30 September 2011, the general trend of the probability forecasts converge to the correct answers. There is, however, much disagreement among the experts. This disagreement persists even near the closing dates of the questions. Unfortunately, the data is not only noisy but also quite sparse. As mentioned before, the forecasters were allowed to update their forecast as long as the questions remained open. Updating, however, was done on a very infrequent basis. An expert gave, on average, only 0.016 forecasts per day, and so a typical expert made a forecast about every 62 days. As a result the average response rate was around 13.5 forecasts per day from a large group of experts.

@@

\section{Discussion}
This paper began with an introduction of a rather unorthodox but nonetheless realistic time-series setting in which probability forecasts are made very infrequently by a heterogeneous group of experts. The resulting data is too sparse to be modeled efficiently with standard time-series methods.  In response to the lack of appropriate modeling procedures, our work introduces an interpretable time-series model that incorporates self-reported expertise to capture a sharp and well-calibrated estimate of the crowd belief. The model is estimated in two-steps: The first step, known as the \textit{sampling step}, samples constrained versions of the model parameters via a Gibbs sampler that only makes use of standard distributions (see Appendix \ref{appendix} for technical details). The second step, known as the \textit{calibration step}, involves a one-dimensional optimization procedure that transforms the constrained parameter values to their unconstrained counterparts.

@@

\subsection{Limitations and Directions for Future Research}
This paper developed a time-series procedure that addresses the main challenges in modeling sparse probability forecasting data. The introduced model was kept parsimonious and can be viewed as a basic template that is amenable to many future extensions. For instance, most of the model parameters were assumed to remain constant over time. It is intuitively reasonable, however, that the behavior of the model parameters is different earlier as opposed to later on in the question.  For instance, the level of disagreement, $\sigma^2_k$, among the experts can be expected to decrease towards the final time point when the the question finally resolves. This hypothesis could be investigated by letting $\sigma^2_{t,k}$ depend on the previous term $\sigma^2_{t-1,k}$.

If personal information on the experts were inquired during the data collection process, it is possible to study the forecasting behavior of different kinds of experts. For instance, in this paper the multiplicative bias was modeled as a function of self-reported expertise. The model, however, is by no means restricted to the study of self-reported expertise. Parameter dependencies could be constructed based on many other expert characteristic, such as gender, education, or specialty, to produce a range of novel insight on the forecasting behavior across different groups of experts. Furthermore, expert characteristics could be interacted with question types, such as economic, domestic, or international, to learn about the kind of experts who generally perform well on certain types of questions. This would be of interest to the decision-maker who could use the information as a basis for consulting only a high-performing subset of the available experts.
